---
title: "Software installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Software installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installing `INLA` and `inlabru`

Due to the work involved in building the binaries for the INLA package C software for different architectures,
the INLA package is not on CRAN, but it can be installed from its own distribution
repository. The easiest approach is to add the repository to your local R options,
e.g. in `~/.Rprofile`,
so that `install.packages("INLA")` can work as usual. Alternatively, the same repository
information can be supplied via `install.packages("INLA", repos = ...)`.

In `~/.Rprofile`, add the following lines, or run the code in your R session:
```{r,eval=FALSE}
local({
  r <- c(INLA = "https://inla.r-inla-download.org/R/testing",
         CRAN = "https://cloud.r-project.org/",
         inlabru_universe = "https://inlabru-org.r-universe.dev")
  options(repos = r)
})
```
Note: if you have installed `INLA` from the `"stable"` repository instead of
`"testing"`, you may need to upgrade to the `"testing"` version,
which typically contains bug fixes and improvements, and is the version `inlabru` is tested against.

The third repository is optional, and gives you easy access to the development
versions of `inlabru` (`r packageVersion("inlabru")` or later) and `fmesher`
(`r packageVersion("fmesher")` or later, which is a package dependency of
`INLA` and `inlabru`).
Most of the tutorials in this course will work with the CRAN versions of `inlabru` and `fmesher`,
`2.12.0`, but installing the development version from `r-universe.dev` is recommended,
and required for a few new features and contains bug fixes, see
[the inlabru changelog](https://inlabru-org.github.io/inlabru/news/).

To install the packages, including optional dependencies many of which are
required for data analysis, run the following code in your R session:
```{r,eval=FALSE}
install.packages(c("INLA", "inlabru"), dependencies = TRUE)
```
The warning about the unavailable `HKprocess` package can safely be ignored.
`INLA` also suggests two Bioconductor packages, `graph` and `Rgraphviz`, which are not
essential for this course, but can be installed with
```{r,eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install(c("graph", "Rgraphviz"), dep = TRUE)
```

## Installation check

You can check that `INLA` is correctly installed by running
```{r}
df <- data.frame(y = rnorm(100) + 10)
fit <- INLA::inla(
  y ~ 1,
  data = df
)
summary(fit)
```
If the simple `inla()` call fails with a crash, you may need to install different
`inla` binaries for your hardware/software combination,
with `INLA::inla.binary.install()`.

When `inla()` works, you can check that `inlabru` is installed correctly by running
the same model in `inlabru`:
```{r}
fit <- inlabru::bru(
  y ~ Intercept(1, prec.linear = exp(-7)),
  data = df
)
summary(fit)
```
